---
title: "Getting Started with BCFM: A Complete Workflow"
author: "Allison Tegge"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BCFM: A Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

library(BCFM)
library(dplyr)    
library(tidyr)    
library(ggplot2)   
```

### Introduction
This vignette demonstrates a complete workflow for using the BCFM (Bayesian Covariance Factor Model) package. We'll use a simulated dataset that comes with the package to illustrate:

- Data preparation and exploration
- Model selection across different numbers of groups and factors
- Fitting the optimal model
- Comprehensive visualization of results

### Load and Explore the Data
For this tutorial, we use a simulated dataset included in the BCFM package. The dataset contains 200 observations, 20 variables, and 5 time points, with a known structure of 4 groups and 3 latent factors.

```{r simulated-data-exploration}
# Load the simulated dataset
data("sim.data", package = "BCFM")

# Examine the structure
str(sim.data)

# Extract dimensions
n_obs <- dim(sim.data$data)[1]      # 200 observations
n_vars <- dim(sim.data$data)[2]     # 20 variables
n_times <- dim(sim.data$data)[3]    # 5 time points

cat("Dataset dimensions:\n")
cat("  Observations:", n_obs, "\n")
cat("  Variables:", n_vars, "\n")
cat("  Time points:", n_times, "\n")
```

### Model Selection
The first step is to identify the optimal number of groups (clusters) and latent factors. We use BCFM.model.selection() to fit models across a grid of possible configurations and compare them using BIC.

```{r model-selection-full, eval=FALSE}
# Full model selection (computationally intensive)
# This code shows the complete workflow but is not run in the vignette
cluster.vars <- paste0("Var", 1:n_vars)

# Create output directory for results
output_dir <- "~/BCFM_analysis"

BCFM.model.selection(
  data = sim.data$data,
  cluster.vars = cluster.vars,
  grouplist = 2:4,        # Try 2, 3, and 4 groups
  factorlist = 2:4,       # Try 2, 3, and 4 factors
  n.iter = 50000,         # Number of MCMC iterations
  burnin = 10000,         # Burnin for BIC calculation
  every = 1000,           # Progress update frequency
  output_dir = output_dir # Specify where to save results
)
```

This will create multiple result files and a `BIC.Rdata` file containing the BIC matrix for model comparison.
*Note on Output Directory:* The output_dir parameter specifies where model results will be saved. This ensures you have write permissions and can easily organize results from multiple analyses. If not specified, results are saved to the current working directory.

```{r examine-bic, eval=FALSE}
# After running model selection, load BIC results from the output directory
load(file.path(output_dir, "BIC.Rdata"))
print(BIC.matrix)
ggplot.BIC(BIC.matrix, factor_list = 2:4, group_list = 2:4)


# Example output:
#       2    3    4
# 2  1234 1156 1189
# 3  1190 1087 1098
# 4  1201 1102 1056  # <- Minimum BIC at 4 groups, 4 factors

# Find optimal configuration
optimal <- which(BIC.matrix == min(BIC.matrix, na.rm = TRUE), arr.ind = TRUE)
cat("Optimal number of groups:", optimal[1], "\n")
cat("Optimal number of factors:", optimal[2], "\n")
```

### Load the Fitted Model
For this demonstration, we'll load a pre-fitted model with 4 groups and 4 factors:

```{r load-results}
# Load pre-computed example results
data("SDresult", package = "BCFM")

# The results are stored in SDresult$Result
str(SDresult$Result)
```

### Visualization
The BCFM package provides comprehensive visualization functions to explore your results.

#### 1. Latent Profiles
Visualize the mean profiles for each cluster:

```{r latent-profiles-plot}
ggplot.latent.profiles(Gibbs = SDresult$Result)
```

#### 2. Factor Loadings Matrix

The factor loadings matrix B shows the relationship between observed variables and latent factors. Each column represents a latent factor, and each row represents a variable. Higher absolute values indicate stronger associations.

```{r factor-loadings-matrix}

# Check dimensions and calculate posterior mean of B (after burnin)
n_iter <- dim(SDresult$Result$B)[1]
burnin <- min(100, floor(n_iter * 0.2))  # Use 20% as burnin or 100, whichever is smaller

cat("Total iterations:", n_iter, "\n")
cat("Using burnin:", burnin, "\n")

# Calculate posterior mean after burnin
B.matrix <- apply(SDresult$Result$B[burnin:n_iter, , ], MARGIN = c(2, 3), FUN = mean)
B.matrix <- as.data.frame(B.matrix)

# Set variable names
rownames(B.matrix) <- paste0("Var", 1:nrow(B.matrix))

# Set informative column names for factors
colnames(B.matrix) <- paste0("Factor ", 1:ncol(B.matrix))

# Print the factor loadings matrix
knitr::kable(round(B.matrix, 3), 
             caption = "Posterior Mean Factor Loadings",
             align = 'c')
```

#### 3. Loading Matrix (B) - Credible Intervals
Examine the factor loadings with credible intervals:

```{r B-credible-intervals-plot}
ggplot.B.CI(SDresult$Result)
```
#### 4. Loading Matrix (B) - Trace Plots
Check MCMC convergence for factor loadings:

```{r trace-plot}
# All variables
ggplot.B.trace(Gibbs = SDresult$Result, factor.num = NULL)
# Specific variable (e.g., variable 1)
ggplot.B.trace(Gibbs = SDresult$Result, factor.num = 1)
```

#### 5. Cluster Means (Mu) - Density Plots
Explore the posterior distributions of cluster means:
```{r mu-density-plot}
ggplot.mu.density(Gibbs = SDresult$Result, add.legend = TRUE)
```

#### 6. Covariance Structure (Omega) - Density Plots
Visualize the within-cluster covariance structure:
```{r omega-density-plot}
# Plot Omega for first cluster
ggplot.omega.density(Gibbs = SDresult$Result, group.select = 1, 
                     show.offdiag = FALSE)

# Plot Omega for second cluster
ggplot.omega.density(Gibbs = SDresult$Result, group.select = 2, 
                     show.offdiag = FALSE)
```

#### 7.  Cluster Probabilities - Density Plots
Examine the posterior distributions of cluster membership probabilities:
```{r probs-density-plot}
ggplot.probs.density(Gibbs = SDresult$Result)
```

#### 8. Cluster Probabilities - Trace Plots
Check convergence of cluster probabilities:

```{r probs-trace-plot}
ggplot.probs.trace(Gibbs = SDresult$Result)
```

#### 9. Error Variances (Sigma²) - Credible Intervals
Visualize the idiosyncratic error variances:
```{r sigma2-credible-intervals-plot}
ggplot.sigma2.CI(Gibbs = SDresult$Result)
```

#### 10. Factor Variances (Tau) - Credible Intervals
Examine the latent factor variances:
```{r tau-credible-intervals-plot}
ggplot.tau.CI(Gibbs = SDresult$Result)
```

#### 11. Variance Decomposition
Understand how variance is partitioned across factors:
```{r variance-plot}
# Use first time point: 200 x 20
data_t1 <- sim.data$data[, , 1]
ggplot.variability(data = data_t1, nfactors = 4)
```

#### 12. Cluster Assignments Heatmap
Visualize the cluster membership assignments over MCMC iterations:
```{r heatmap-plot}
ggplot.Zit.heatmap(Gibbs = SDresult$Result)
```

### Interpreting Results

#### Cluster Assignments
The heatmap shows how observations are assigned to clusters. Stable assignments (consistent colors across iterations) indicate good convergence and clear cluster structure.

#### Factor Loadings
The loading matrix B shows which variables are associated with each latent factor. Large credible intervals that include zero suggest weak associations.

#### Cluster Characteristics
The latent profiles plot reveals the distinctive patterns of each cluster across variables, helping to interpret what makes each cluster unique.

#### Accessing Model Components
You can directly access various components of the fitted model:
```{r model-components}
# Cluster assignments (most likely cluster for each observation)
cluster_assignments <- SDresult$Result$Z

# Loading matrix
loadings <- SDresult$Result$B

# Cluster means
cluster_means <- SDresult$Result$mu

# Cluster probabilities
cluster_probs <- SDresult$Result$probs

# Factor scores
factor_scores <- SDresult$Result$X

# Show dimensions
cat("Dimensions of key components:\n")
cat("  B (loadings):", dim(loadings), "\n")
cat("  mu (means):", dim(cluster_means), "\n")
cat("  X (scores):", dim(factor_scores), "\n")
```

### Session Info
```{r}
sessionInfo()
```
