---
title: "Getting Started with BCFM: A Complete Workflow"
author: "Allison Tegge"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with BCFM: A Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)

library(BCFM)
library(dplyr)    # For %>% pipe operator
library(tidyr)    # For pivot_longer
library(ggplot2)  # For plotting
```

### Introduction
This vignette demonstrates a complete workflow for using the BCFM (Bayesian Covariance Factor Model) package. We'll use a simulated dataset that comes with the package to illustrate:

- Data preparation and exploration
- Model selection across different numbers of groups and factors
- Fitting the optimal model
- Comprehensive visualization of results

### Load and Explore the Data
For this tutorial, we use a simulated dataset included in the BCFM package. The dataset contains 200 observations, 20 variables, and 5 time points, with a known structure of 4 groups and 3 latent factors.

```{r}
# Load the simulated dataset
data("sim.data", package = "BCFM")

# Examine the structure
str(sim.data)

# Extract dimensions
n_obs <- dim(sim.data$data)[1]      # 200 observations
n_vars <- dim(sim.data$data)[2]     # 20 variables
n_times <- dim(sim.data$data)[3]    # 5 time points

cat("Dataset dimensions:\n")
cat("  Observations:", n_obs, "\n")
cat("  Variables:", n_vars, "\n")
cat("  Time points:", n_times, "\n")
```

### Model Selection
The first step is to identify the optimal number of groups (clusters) and latent factors. We use BCFM.model.selection() to fit models across a grid of possible configurations and compare them using BIC.

```{r}
# Specify which variables to use for clustering (all 20 variables)
cluster.vars <- paste0("Var", 1:n_vars)

# Run model selection over a grid
# We'll test different numbers of groups and factors
BCFM.model.selection(
  data = sim.data$data,
  cluster.vars = cluster.vars,
  grouplist = 2:4,        # Try 2, 3, and 4 groups
  factorlist = 2:4,       # Try 2, 3, and 4 factors
  n.iter = 10000,         # Number of MCMC iterations
  burnin = 5000,          # Burnin for BIC calculation
  every = 2000            # Progress update frequency
)
```

This will create multiple result files and a `BIC.Rdata` file containing the BIC matrix for model comparison.

```{r}
# Load BIC results
load("BIC.Rdata")
print(BIC.matrix)

# Find optimal configuration
optimal <- which(BIC.matrix == min(BIC.matrix, na.rm = TRUE), arr.ind = TRUE)
cat("Optimal number of groups:", optimal[1], "\n")
cat("Optimal number of factors:", optimal[2], "\n")
```

### Load the Fitted Model
For this demonstration, we'll load a pre-fitted model with 4 groups and 4 factors:

```{r}
# Load pre-computed example results
data("SDresult", package = "BCFM")

# The results are stored in SDresult$Result
str(SDresult$Result)
```

### Visualization
The BCFM package provides comprehensive visualization functions to explore your results.

#### 1. Latent Profiles
Visualize the mean profiles for each cluster:

```{r}
ggplot.latent.profiles(Gibbs = SDresult$Result)

```

#### 2. Factor Loadings Matrix

The factor loadings matrix B shows the relationship between observed variables and latent factors. Each column represents a latent factor, and each row represents a variable. Higher absolute values indicate stronger associations.

```{r factor-loadings-prep, eval=FALSE}
# Calculate posterior mean of B (after burnin)
B.matrix <- apply(SDresult$Result$B[100:5000, , ], MARGIN = c(2, 3), FUN = mean)
B.matrix <- as.data.frame(B.matrix)

# Set variable names
rownames(B.matrix) <- paste0("Var", 1:nrow(B.matrix))

# Set informative column names for factors
colnames(B.matrix) <- paste0("Factor ", 1:ncol(B.matrix))

# Print the factor loadings matrix
knitr::kable(round(B.matrix, 3), 
             caption = "Posterior Mean Factor Loadings",
             align = 'c')
```

#### 3. Loading Matrix (B) - Credible Intervals
Examine the factor loadings with credible intervals:

```{r}
ggplot.B.CI(SDresult$Result)
```
#### 4. Loading Matrix (B) - Trace Plots
Check MCMC convergence for factor loadings:

```{r}
# All variables
ggplot.B.trace(Gibbs = SDresult$Result, factor.num = NULL)
# Specific variable (e.g., variable 1)
ggplot.B.trace(Gibbs = SDresult$Result, factor.num = 1)
```

#### 5. Cluster Means (Mu) - Density Plots
Explore the posterior distributions of cluster means:
```{r}
ggplot.mu.density(Gibbs = SDresult$Result, add.legend = TRUE)
```

#### 6. Covariance Structure (Omega) - Density Plots
Visualize the within-cluster covariance structure:
```{r}
# Plot Omega for first cluster
ggplot.omega.density(Gibbs = SDresult$Result, group.select = 1, 
                     show.offdiag = FALSE)

# Plot Omega for second cluster
ggplot.omega.density(Gibbs = SDresult$Result, group.select = 2, 
                     show.offdiag = FALSE)
```

#### 7.  Cluster Probabilities - Density Plots
Examine the posterior distributions of cluster membership probabilities:
```{r}
ggplot.probs.density(Gibbs = SDresult$Result)
```

#### 8. Cluster Probabilities - Trace Plots
Check convergence of cluster probabilities:

```{r}
ggplot.probs.trace(Gibbs = SDresult$Result)
```

#### 9. Error Variances (Sigma²) - Credible Intervals
Visualize the idiosyncratic error variances:
```{r}
ggplot.sigma2.CI(Gibbs = SDresult$Result)
```

#### 10. Factor Variances (Tau) - Credible Intervals
Examine the latent factor variances:
```{r}
ggplot.tau.CI(Gibbs = SDresult$Result)
```

#### 11. Variance Decomposition
Understand how variance is partitioned across factors:
```{r}
# Use first time point: 200 x 20
data_t1 <- sim.data$data[, , 1]
ggplot.variability(data = data_t1, nfactors = 4)
```

#### 12. Cluster Assignments Heatmap
Visualize the cluster membership assignments over MCMC iterations:
```{r}
ggplot.Zit.heatmap(Gibbs = SDresult$Result)
```
### Interpreting Results
#### Cluster Assignments
The heatmap shows how observations are assigned to clusters. Stable assignments (consistent colors across iterations) indicate good convergence and clear cluster structure.
#### Factor Loadings
The loading matrix B shows which variables are associated with each latent factor. Large credible intervals that include zero suggest weak associations.
#### Cluster Characteristics
The latent profiles plot reveals the distinctive patterns of each cluster across variables, helping to interpret what makes each cluster unique.
#### Accessing Model Components
You can directly access various components of the fitted model:
```{r}
```{r}
# Cluster assignments (most likely cluster for each observation)
cluster_assignments <- SDresult$Result$Z

# Loading matrix
loadings <- SDresult$Result$B

# Cluster means
cluster_means <- SDresult$Result$mu

# Cluster probabilities
cluster_probs <- SDresult$Result$probs

# Factor scores
factor_scores <- SDresult$Result$X

# Show dimensions
cat("Dimensions of key components:\n")
cat("  B (loadings):", dim(loadings), "\n")
cat("  mu (means):", dim(cluster_means), "\n")
cat("  X (scores):", dim(factor_scores), "\n")
```

### Session Info
```{r}
sessionInfo()
```
